---
title: "Assignment 5 Data Report"
output: html_document
---

``` {r setup, include = FALSE, echo = FALSE}
source("analysis.R")

library(knitr)
```

# Mass shootings in 2018: a data report

Gun violence has quickly become one of the most prominent issues in the US
over the past decade.

``` {r link1, echo = FALSE}
l1_markdown <-
  "[2018 report](https://www.gunviolencearchive.org/reports/mass-shooting)"
```

According to the Gun Violence Archive's (GVA) `r l1_markdown`, there were a 
total of `r sum_shootings` reported incidents. The shootings caused a total of
`r sum_lost` lives lost. The city that was most impacted by shootings in 2018
was `r sum_city` (impact is analyzed by the amount of lives lost in shootings,
however, it is respectfully recognized that some reports may calculate impact by
injuries). The state that was most impacted by shootings in 2018 was 
`r sum_state`. It was also found that the month with the most reported shootings
was `r sum_month`.

## Summary of GVA 2018 Shooting Data

``` {r echo = FALSE}
kable(table_df,
  col.names = c(
    "State",
    "Total Victims",
    "Deaths",
    "Injuries",
    "Death Percentage",
    "Injury Percentage"
  ),
  align = "r"
)
```

Many of the state injury to death ratios favor injuries, but there still are
some states that show the opposite where there are more deaths than injuries.
However, it should be noted that these states usually have lower victim totals,
indicating that there could be a positive correlation between total victims and
the ratio of injuries to deaths.

## Many mass shootings means many unfortunate victims

On `r spfc_date`, a shooting occurred in `r spfc_city_state`. This incident - that
happened at `r spfc_address` - would become one of the most prominent shootings
of the year, totaling up to `r spfc_injr` and `r spfc_lost`. These 
`r spfc_injr + spfc_lost` victims were the result of the worst shooting that 
year according to the amount of reported injuries. The shooting was one of
`r spfc_in_month` shootings that happened in `r spfc_city` that month.

``` {r link2, echo = FALSE}
l2_markdown <- paste0(
  "[here]",
  "(https://www.nytimes.com/2018/06/17/nyregion/",
  "trenton-mass-shooting.html)"
)
```

A link to the report on the shooting incident can be found `r l2_markdown`
(note that data from this report might differ slightly from the given link).

## Incident map

The map below represents the shooting locations according to the 2018 GVA
report. The size of the circle represents the number of victims during the
incident relative to other incidents.

``` {r map, echo = FALSE}
library(leaflet)
library(htmltools)

# map of data
leaflet(data = shootings) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircles(
    lat = ~lat,
    lng = ~long,
    radius = ~ num_killed * 5000,
    stroke = FALSE,
    label = lapply(labels, htmltools::HTML) # label uses html tools to convert
    # char vector 'labels' to HTML
  )
```

As seen from a geographical view, the majority of the larger mass shootings of
2018 happened on the east coast, with a very concentrated amount of incident 
reports in the Maryland/New Jersey/Pennsylvania area. The largest concentration
of incident reports on the west coast are seen in the LA area. However, if we 
take a look at the same data from a different angle, we see a very different 
story.

## Analysis by state

As seen above, states saw varying levels of activity in 2018. The states that
experienced mass shootings had differing numbers of victims, and of those
victims, had differing ratios of casualties to injuries. However, the
geographical perspective was less effective at showing how areas with high
concentration of shooting reports were really effected.

``` {r bar plot, echo = FALSE}
library(ggplot2)

ggplot(data = by_state, aes(
  x = reorder(state, total),
  y = total,
  fill = victim_type
)) +
  geom_col() +
  labs(
    title = "Shooting victims per state (2018)",
    x = "State",
    y = "Total victims"
  ) +
  scale_fill_brewer(
    name = "Victim types",
    labels = c("Injuries", "Deaths"),
    palette = "Set1",
    direction = -1
  ) +
  coord_flip()
```

A horizontal bar chart was chosen to reflect the number of victims in each 
state. Notice that the chart is also stacked with both number of injuries
and deaths for each state. This allows viewers to more easily compare the
number of victims to each state than the map above. For example, the map shows
a high concentration of dots in the north eastern area of the US, however, it's
obvious according to the bar chart that both California and Florida had the 
most victims. Note: this may not necessarily mean that there were more 
shootings in those states than the north eastern states; another chart should
be made to explicitly compare shooting incident density. Although, the map may
not have been extremely effective, it still managed to depict some of the data
correctly. Pennsylvania, a north eastern state, had the most victims per state
of those under 100 victims.




